---
title: "Timeseries"
author: "Julianna Renzi"
date: "1/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(lubridate)
library(tsibble) # time series tibbles
library(feasts) # wrangling and working with time series data
library(slider) # finding moving averages
```

## Read in the data

```{r}
toolik <- read_csv(here("data", "toolikweather.csv")) # date is a character initially
  


```


### Convert this into a tsibble

```{r}
toolik_ts <- toolik %>% 
  mutate(date = lubridate::mdy(date)) %>% # first convert it into a date class that R understands
  as_tsibble(key = NULL, #  key would be stations or different sites if you had them
             index = date) 


ggplot(data = toolik_ts, aes(x = date, y = mean_airtemp)) + 
  geom_line() + 
  theme_classic()

```

## Use `index_by()` function to aggregate time series information by intervals we specify (timeseries equivalent of group_by())

First calculate average monthly temperature

```{r}
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% # this is applied to both year and month (yearmonth) and to the whole specified vector (.)
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))

ggplot(data = toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line() + 
  facet_wrap(~month(yr_mo, label = TRUE))
# COOL! Can still pull out year/month from yr_mo using lubridate functions
```

Aggregate by week:

```{r}
toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% # yearweek etc. are built in syntax--look them up
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE),
    min_airtemp = min(mean_airtemp))

head(toolik_weekly)
```
## Filtering time series

This is the time series equivalent of filtering. Use `filter_index()` to specify ranges of time series to keep or exclude.

```{r}
# filter between start of June 2000 to end of Oct 2001
toolik_ts %>% 
  filter_index("2000-06" ~ "2001-10")
```
Can filter at a higher resolution, too

```{r}
# Ap 10 to May 5
toolik_ts %>% 
  filter_index("2006-04-10" ~ "2006-05-15")

# check out different filter_index() options
```

## Seasonplots

```{r}
toolik_ts %>% 
  filter(year(date) > 2014) %>% # this is the same as filter_index with 2015 ~ . (dot would be end of the dataset)
  gg_season(y = mean_airtemp)

# can also do it monthly
toolik_month %>% 
  gg_season(y = monthly_mean_temp) + # probably want to make it a continuous color scale since the numbers mean things
  theme_light()

```


## Seasonal subseries plot

Allows you to break up subseries (like facet_wrapping, but a bit easier)

```{r}
toolik_month %>% 
  gg_subseries(monthly_mean_temp) # like what we did before, but a little cleaner and has month across all months

```

## Find the moving average using `slider` package

```{r}
set.seed(2021)
test <- rnorm(n = 100, mean = 40, sd = 10)

# see different windows
slide(test, ~.x, .before = 2, .after = 2) # across all values of the vector, creates a window of 5 (2 before, 2 after)
# first ones don't have all 5 because don't have 2 before/2 after
# this pulls all the values and puts it in a list--want to convert it to a numeric vector to use it for graphing


```
Rolling (moving average) for windows of size 15

```{r}
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp, # variable to use
                                   mean, # function to apply
                                   .before = 7, # window size (doesn't include center value). In this case it's days
                                   .after = 7)))


# plot it
ggplot() +
  geom_line(data = toolik_ts, aes(x = date,
                                  y = mean_airtemp), size = 0.2, color = "grey") + # original data
  # add moving average
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d)) + # slightly smoother (larger windows will give smoother results)
  theme_minimal()
```

## Create autocorrelation function (ACF)

How observations are correlated with prior observations within the same variable

```{r}
# start with days
toolik_ts %>% 
  ACF(mean_airtemp) %>% # part of feasts package--this gives you the values
  autoplot() # this is if the lag is days
# see observations are pretty highly correlated (i.e. days within a month of each other are pretty correlated)
# BUT seasonality is probably better represented with months

toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot() # see clear and reflected seasonality now in the ACF (lag on x-axis is in month and see observations 12 months apart are strongly correlated whereas those 6 months apart are strongly negatively correlated)
# also see many years back the correlation is still very strong (not a lot of dampening happening)
```

## Time series decomposition by STL

This is a type better than classical decomposition: STL (seasonal trend and loess). Similar: have seasonal, trend, and remainder components

```{r}
toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf))) # Inf window means the same window as X (month)

# pull seasonal trend and remainder from model
components(toolik_dec) %>% 
  autoplot()
# first panel is original, second is trend, third is seasonality, fourth is residual
# NEED TO LOOK AT THE SCALES! First panel has a range of ~40, trend has a very small y-axis (4, or 10% of change we see in the total data)--probably not that important
# seasonality on the other hand is on a large scale--seasonality is a huge part of this time series (range of like 30!)
```

